---
title: "Divison of Undergraduate Education"
author: "Dominik Glandorf"
date: "2022-11-11"
output: pdf_document
---

```{r read data, include=F}
Sys.setenv(RSTUDIO_PANDOC = "C:/Program Files/RStudio/bin/quarto/bin/tools")

# set wd to source file directory
source('read_data.R')

students = get_student_sub()
bg_sub = get_student_background_data(ids = students$mellon_id)
terms = get_term_data(ids = students$mellon_id)
courses = get_course_data(ids = students$mellon_id)
term_features = get_term_features()
```

# Division of Undergraduate Education
```{r, echo=F}
missing_school = mean(is.na(terms$major_school_name_1))
```
Students are assigned to one or more schools for each term (missing for `r round(missing_school*100,1)`% of terms). These are the schools occurring in the data:
```{r, echo=F}
unique(terms$major_school_name_1)
```
```{r, echo=F}
DUE = "Div of Undergraduate Education"
is_term_school_DUE = (terms$major_school_name_1 == DUE)
terms_DUE = mean(is_term_school_DUE, na.rm=T)

student_ever_DUE = aggregate(major_school_name_1 ~ mellon_id, terms, FUN=function(x) DUE %in% x)
mean_ever_DUE = mean(student_ever_DUE$major_school_name_1)
student_ever_DUE_2 = aggregate(cbind(major_school_name_1, major_school_name_2) ~ mellon_id, terms, FUN=function(x) DUE %in% x)
mean(student_ever_DUE_2$major_school_name_2) # never occurs as second school

student_always_DUE = aggregate(major_school_name_1 ~ mellon_id, terms, FUN=function(x) DUE %in% unique(x) & length(unique(x)) == 1)
mean_always_DUE = mean(student_always_DUE$major_school_name_1)

student_first_DUE = aggregate(major_school_name_1 ~ mellon_id, terms, FUN=function(x) DUE == x[1])
mean_first_DUE = mean(student_first_DUE$major_school_name_1)
```
`r round(mean_ever_DUE*100,1)`% of students are part of it at least once, overall in `r round(terms_DUE*100,2)`% of terms. However, `r round(mean_always_DUE*100,2)`% of students have been part of DUE in all of their terms whereas `r round(mean_first_DUE*100,2)`% have been in their first term.

```{r, fig.height=3, fig.width=8, echo=F}
ggplot(term_features, aes(fill=(school_1 == DUE), x=term_num)) + 
    geom_bar(position="stack", stat="count") +
    labs(title = "Distribution of DUE over terms",
       x = "Term number",
       y = "Frequency", fill="part of DUE")
```
The share of DUE students decreases over time. Few students (`r round((mean_ever_DUE-mean_first_DUE)*100,1)`%) join DUE after their first term.

```{r, echo=F, fig.height=3, fig.width=8}
student_ever_DUE = merge(student_ever_DUE, students[,c("mellon_id", "dropout", "admitdate")], by="mellon_id")
student_first_DUE = merge(student_first_DUE, students[,c("mellon_id", "dropout", "admitdate")], by="mellon_id")
student_always_DUE = merge(student_always_DUE, students[,c("mellon_id", "dropout", "admitdate")], by="mellon_id")

ggplot(student_ever_DUE, aes(x=admitdate, fill=major_school_name_1))+
  geom_bar(stat="count") +
    labs(title = "Distribution of DUE over admission",
       x = "Admission",
       y = "Frequency", fill="Ever part of DUE")
```
The school is much often assigned for students admitted in later years.

# Dropout
```{r, include=F}
dropout_ever_DUE = mean(student_ever_DUE$dropout[student_ever_DUE$major_school_name_1==TRUE], na.rm=T)
dropout_never_DUE = mean(student_ever_DUE$dropout[student_ever_DUE$major_school_name_1==FALSE], na.rm=T)

dropout_first_DUE = mean(student_first_DUE$dropout[student_first_DUE$major_school_name_1==TRUE], na.rm=T)
dropout_not_first_DUE = mean(student_first_DUE$dropout[student_first_DUE$major_school_name_1==FALSE], na.rm=T)

dropout_always_DUE = mean(student_always_DUE$dropout[student_always_DUE$major_school_name_1==TRUE], na.rm=T)
dropout_not_always_DUE = mean(student_always_DUE$dropout[student_always_DUE$major_school_name_1==FALSE], na.rm=T)
```

`r round(dropout_ever_DUE*100,1)`% of students that have ever been part of DUE drop out, compared to `r round(dropout_never_DUE*100,2)`% among those who have never been.

`r round(dropout_first_DUE*100,1)`% of students that have been part of DUE in first term drop out, compared to `r round(dropout_not_first_DUE*100,2)`% among those who have not been.

`r round(dropout_always_DUE*100,1)`% of students that have always been part of DUE drop out, compared to `r round(dropout_not_always_DUE*100,2)`% among those who have not always been.


```{r, include=F}
# Example transcripts
source("inspect_student.R")
DUE_students = sample(student_ever_DUE$mellon_id[student_ever_DUE$major_school_name_1==TRUE],5)
for (s_id in DUE_students) {
  inspect_student(s_id)
}
```

# How are DUE students different from others?
## Do they study a shorter time at UCI?
```{r}
num_terms = aggregate(term_num ~ mellon_id, term_features, FUN=max)
student_ever_DUE = merge(student_ever_DUE, num_terms, by="mellon_id")
is_DUE = student_ever_DUE$major_school_name_1==TRUE
dropout = student_ever_DUE$dropout==TRUE

terms_DUE_grads = mean(student_ever_DUE$term_num[is_DUE&!dropout], na.rm=T)
terms_non_DUE_grads = mean(student_ever_DUE$term_num[!is_DUE&!dropout], na.rm=T)

terms_DUE_DO = mean(student_ever_DUE$term_num[is_DUE&dropout], na.rm=T)
terms_non_DUE_DO = mean(student_ever_DUE$term_num[!is_DUE&dropout], na.rm=T)
```
Graduates both ever been at DUE or not usually study `r round(terms_DUE_grads,1)` or  `r round(terms_non_DUE_grads,1)` terms. Dropouts ever been at DUE only study `r round(terms_DUE_DO,1)` compared to `r round(terms_non_DUE_DO,1)`.

## Is it the ones that did not declare their major or were unaffiliated in the first term?
```{r, include=F}
term_features_1 = subset(term_features, term_num==1)
is_DUE = term_features_1$school_1==DUE
unaff = mean(term_features_1$major_1=="UNAFFILIATED", na.rm=T)
unaff_DUE = mean(term_features_1$major_1[is_DUE]=="UNAFFILIATED", na.rm=T)
unaff_non_DUE = mean(term_features_1$major_1[!is_DUE]=="UNAFFILIATED", na.rm=T)
undec_DUE = mean(term_features_1$major_1[is_DUE]=="UNDECLARED", na.rm=T)
undec_non_DUE = mean(term_features_1$major_1[!is_DUE]=="UNDECLARED", na.rm=T)
```
Almost no students are unaffiliated in the first term (`r round(unaff*100,2)`%), neither DUE nor non-DUE students.

Among the DUE students `r round(undec_DUE*100,1)`% did not declare a major in the first term. However, among the remainder this is the case for `r round(undec_non_DUE*100,1)`%. 

## Is it international students?
```{r, include=F}
student_ever_DUE = merge(student_ever_DUE, bg_sub[,c('mellon_id','int_student')], by="mellon_id")
is_DUE = student_ever_DUE$major_school_name_1==TRUE
dropout = student_ever_DUE$dropout==TRUE
int_student = student_ever_DUE$int_student==TRUE

int_DUE = mean(student_ever_DUE$int_student[is_DUE], na.rm=T)
int_non_DUE = mean(student_ever_DUE$int_student[!is_DUE], na.rm=T)
```
The ratio of international students only differs slightly between ever DUE and never DUE students (`r round(int_DUE*100,1)`% vs `r round(int_non_DUE*100,1)`%).

## Do they differ demographically?
```{r}
student_ever_DUE = merge(student_ever_DUE, bg_sub[,c('mellon_id','ethnicity','female')], by="mellon_id")

ggplot(data = student_ever_DUE, aes(x = ethnicity)) +
  geom_bar(data=subset(student_ever_DUE,major_school_name_1 == T),aes(y = (..count..)/sum(..count..)),fill = "red", alpha = 0.4,width=0.5,just=0.0) +
  geom_bar(data=subset(student_ever_DUE,major_school_name_1 == F),aes(y = (..count..)/sum(..count..)),fill = "darkgreen", alpha = 0.4, width=0.5, just=1) +
  theme_minimal() +
  labs(title = "Distribution of ethnicity (DUE is red)",
       x = "Ethnicity",
       y = "Share") +
  theme(text = element_text(size = 16)) +
  scale_x_discrete(label=abbreviate)
```
```{r}
ggplot(data = student_ever_DUE, aes(x = female)) +
  geom_bar(data=subset(student_ever_DUE,major_school_name_1 == T),aes(y = (..count..)/sum(..count..)),fill = "red", alpha = 0.4,width=0.5,just=0.0) +
  geom_bar(data=subset(student_ever_DUE,major_school_name_1 == F),aes(y = (..count..)/sum(..count..)),fill = "darkgreen", alpha = 0.4, width=0.5, just=1) +
  theme_minimal() +
  labs(title = "Distribution of gender (DUE is red)",
       x = "Is female",
       y = "Share") +
  theme(text = element_text(size = 16)) +
  scale_x_discrete(label=abbreviate)
```

## Do they perform differently in pre uni scores?

```{r}
student_ever_DUE = merge(student_ever_DUE, students[,c('mellon_id','hs_gpa','toefl_score','uc_total_score')], by="mellon_id")
```

### Highschool GPA
```{r}
ggplot(data = student_ever_DUE, aes(x = round(hs_gpa*5)/5)) +
  geom_bar(data=subset(student_ever_DUE,major_school_name_1 == T),aes(y = (..count..)/sum(..count..)),fill = "red", alpha = 0.4) +
  geom_bar(data=subset(student_ever_DUE,major_school_name_1 == F),aes(y = (..count..)/sum(..count..)),fill = "darkgreen", alpha = 0.4) +
  theme_minimal() +
  labs(title = "Distribution of highschool GPA (DUE is red)",
       x = "GPA",
       y = "Share") +
  theme(text = element_text(size = 16))
```
### UC total score
```{r}
ggplot(data = student_ever_DUE, aes(x = round(uc_total_score/10)*10)) +
  geom_bar(data=subset(student_ever_DUE,major_school_name_1 == T),aes(y = (..count..)/sum(..count..)),fill = "red", alpha = 0.4) +
  geom_bar(data=subset(student_ever_DUE,major_school_name_1 == F),aes(y = (..count..)/sum(..count..)),fill = "darkgreen", alpha = 0.4) +
  theme_minimal() +
  labs(title = "Distribution of UC total score (DUE is red)",
       x = "Total score",
       y = "Share") +
  theme(text = element_text(size = 16))
```
### TOEFL score
```{r}
ggplot(data = student_ever_DUE, aes(x = round(toefl_score/5)*5)) +
  geom_bar(data=subset(student_ever_DUE,major_school_name_1 == T),aes(y = (..count..)/sum(..count..)),fill = "red", alpha = 0.4) +
  geom_bar(data=subset(student_ever_DUE,major_school_name_1 == F),aes(y = (..count..)/sum(..count..)),fill = "darkgreen", alpha = 0.4) +
  theme_minimal() +
  labs(title = "Distribution of TOEFL score (DUE is red)",
       x = "TOEFL score",
       y = "Share") +
  theme(text = element_text(size = 16))
```
## Do they perform differently in first term?
```{r}
student_gpas = aggregate(gpa_term ~ mellon_id, terms, FUN=mean)

student_ever_DUE = merge(student_ever_DUE, student_gpas[,c('mellon_id','gpa_term')], by="mellon_id")

ggplot(data = student_ever_DUE, aes(x = round(gpa_term*10)/10)) +
  geom_bar(data=subset(student_ever_DUE,major_school_name_1 == T),aes(y = (..count..)/sum(..count..)),fill = "red", alpha = 0.4) +
  geom_bar(data=subset(student_ever_DUE,major_school_name_1 == F),aes(y = (..count..)/sum(..count..)),fill = "darkgreen", alpha = 0.4) +
  theme_minimal() +
  labs(title = "Distribution of average GPA (DUE is red)",
       x = "GPA",
       y = "Share") +
  theme(text = element_text(size = 16))
```

