---
title: "Missing Analysis"
author: "Dominik Glandorf"
date: "2022-11-22"
output: pdf_document
---

# Background variables
```{r, include=F}
student_vars <- read.csv("~/EWS/Variable_request_EWS_Student.csv")
requested_vars = student_vars$Variable.name
source('read_data.R')
students = get_student_sub()
bg = get_student_background_data(students$mellon_id)
available_vars = names(bg)
expected_available = requested_vars[requested_vars %in% available_vars]
missing_in_data = requested_vars[!requested_vars %in% available_vars]
unexpected_in_data = available_vars[!available_vars %in% requested_vars]
types = sapply(bg, class)
```

The spreadsheet knows `r length(requested_vars)` variables. The corresponding file contains `r length(available_vars)` columns. The expected variables `r paste(missing_in_data, collapse=",")` are not in the dataset. Therefore exist variables `r paste(unexpected_in_data, collapse=",")`.
```{r, include=F}
bg_vars = data.frame(name=requested_vars, requested=T)
bg_vars$in_data = bg_vars$name %in% available_vars
extra_vars = data.frame(name=unexpected_in_data, in_data=T, requested=F)
bg_vars = rbind(bg_vars, extra_vars)
bg_vars = column_to_rownames(bg_vars, "name")
bg_vars$use_raw = NA
bg_vars$use_feature = NA
bg_vars$exclude = NA
bg_vars$num_unique = NA
bg_vars$examples = NA
bg_vars$type=NA
for (var in c(available_vars, unexpected_in_data)) {
  bg_vars[var,'type'] = types[var]
}
bg_vars$total_missing = NA
admitdates = unique(bg$admitdate)
for (a in admitdates[order(admitdates)]) {
  bg_vars[paste0('missing_', a)] = NA
}
```


## Available variables
```{r, echo=F, comment=NA}
predictors=c()
blacklist="mellon|student_sid|group|desc|cohort" # should not be used as a predictor
is_derived_from="sat|act|ap_|birth" # is used in feature engineering
for (var in rownames(subset(bg_vars, in_data))) {
  print(paste0('Variable: ', var, ", type: ", types[var], ', requested: ', if(bg_vars[var,'requested'])'yes'else'no'), quote=F)
  
  unique_values = unique(bg[,var])
  num_unique = nrow(unique_values)
  bg_vars[var,'num_unique'] = num_unique
  
  first_values = unique_values[1:min(num_unique, 5),var]
  bg_vars[var,'examples'] = paste0(pull(first_values,var),collapse=", ")
  
  print(paste0('Values (', num_unique, ' unique): ', paste0(pull(first_values,var),collapse=", "), if(num_unique > 5) ', ...' else ''),quote=FALSE)
  
  missing = mean(is.na(bg[,var]))
  bg_vars[var,'total_missing'] = missing
  print(paste0("Missing: ", round(missing*100, 1), "%"), quote=F)
  
  missing_cohorts = aggregate(is.na(bg[,var]), by=list(bg$admitdate), FUN=mean)
  for (a in admitdates) {
    bg_vars[var,paste0('missing_', a)] = missing_cohorts[missing_cohorts$Group.1==a,var]
  }
  if (missing > 0.01 && missing < 0.99) {
    max = which.max(missing_cohorts[,var])
    min = which.min(missing_cohorts[,var])
    if (missing_cohorts[max,var]-missing_cohorts[min,var]>.5) {
      print(missing_cohorts)
    } else {
      print(paste0("Most missing: ", missing_cohorts[max,"Group.1"], " ", round(missing_cohorts[max,var]*100, 1), "%", ", Least missing: ", missing_cohorts[min,"Group.1"], " ", round(missing_cohorts[min,var]*100, 1), "%"), quote=F)
    }
  }
  if (var != "city_residence_app") {
     print(ggplot(bg, aes(x=.data[[var]])) +
    geom_bar() +
    labs(x=var, y="Frequency"))
  }
 
  
  # print statusses
  if (grepl( blacklist , var)) {
    bg_vars[var,'exclude'] = T 
    print("should not be used as a predictor",quote=F)
  } else {
    bg_vars[var,'exclude'] = F
  } 
  if (grepl( is_derived_from , var)) {
    bg_vars[var,'use_feature'] = T 
    print("is used in feature engineering",quote=F)
  } else {
    bg_vars[var,'use_feature'] = F
  } 
  # filter for variables that contain enough data, are continuous or a factor
  if(missing < 0.5 &
     (types[var] %in% c("numeric,integer") | num_unique <= 10)) {
    predictors = c(predictors, var)
    # TODO: decide if to use raw
  }
  

  print("_________________________________________",quote=F)
}
```
These are the predictors that remain when filtering for less than 50% missing, our blacklist and variables that need to be somehow transformed first: 
```{r, echo=F, comment=NA}
predictors
```
```{r}
write.csv2(bg_vars, 'bg_vars.csv', )
```


# Student by term
```{r, include=F}
term_vars <- read.csv("~/EWS/Variable_request_EWS_StudentTerm.csv")
requested_vars = term_vars$Variable.Name

terms = get_term_data(students$mellon_id)
available_vars = names(terms)
expected_available = requested_vars[requested_vars %in% available_vars]
missing_in_data = requested_vars[!requested_vars %in% available_vars]
unexpected_in_data = available_vars[!available_vars %in% requested_vars]
types = sapply(terms, class)
```

The spreadsheet knows `r length(requested_vars)` variables. The corresponding file contains `r length(available_vars)` columns. The expected variables `r paste(missing_in_data, collapse=",")` are not in the dataset. Therefore exist variables `r paste(unexpected_in_data, collapse=",")`.
```{r, include=F}
term_vars = data.frame(name=requested_vars, requested=T)
term_vars$in_data = term_vars$name %in% available_vars
extra_vars = data.frame(name=unexpected_in_data, in_data=T, requested=F)
term_vars = rbind(term_vars, extra_vars)
term_vars = column_to_rownames(term_vars, "name")
term_vars$use_raw = NA
term_vars$use_feature = NA
term_vars$exclude = NA
term_vars$num_unique = NA
term_vars$examples = NA
term_vars$type=NA
for (var in c(available_vars, unexpected_in_data)) {
  term_vars[var,'type'] = types[var]
}
term_vars$total_missing = NA
admitdates = unique(bg$admitdate)
for (a in admitdates[order(admitdates)]) {
  term_vars[paste0('missing_', a)] = NA
}
```
```{r}
# merge admitdate to terms
terms = merge(terms, bg[,c('mellon_id','admitdate')], by="mellon_id")
```

## Available variables
```{r, echo=F, comment=NA}
predictors=c()
blacklist="mellon_id|group_" # should not be used as a predictor
is_derived_from="___" # is used in feature engineering
for (var in rownames(subset(term_vars, in_data))) {
  print(paste0('Variable: ', var, ", type: ", types[var], ', requested: ', if(term_vars[var,'requested'])'yes'else'no'), quote=F)
  
  unique_values = unique(terms[,var])
  num_unique = length(unique_values)
  term_vars[var,'num_unique'] = num_unique
  
  first_values = unique_values[1:min(num_unique, 5)]
  term_vars[var,'examples'] = paste0(first_values,collapse=", ")
  
  print(paste0('Values (', num_unique, ' unique): ', paste0(first_values,collapse=", "), if(num_unique > 5) ', ...' else ''),quote=FALSE)
  
  missing = mean(is.na(terms[,var]))
  term_vars[var,'total_missing'] = missing
  print(paste0("Missing: ", round(missing*100, 1), "%"), quote=F)
  
  missing_cohorts = aggregate(is.na(terms[,var]), by=list(terms$admitdate), FUN=mean)
  for (a in admitdates) {
    term_vars[var,paste0('missing_', a)] = missing_cohorts[missing_cohorts$Group.1==a,'x']
  }
  if (missing > 0.01 && missing < 0.99) {
    max = which.max(missing_cohorts[,'x'])
    min = which.min(missing_cohorts[,'x'])
    if (missing_cohorts[max,'x']-missing_cohorts[min,'x']>.5) {
      print(missing_cohorts)
    } else {
      print(paste0("Most missing: ", missing_cohorts[max,"Group.1"], " ", round(missing_cohorts[max,'x']*100, 1), "%", ", Least missing: ", missing_cohorts[min,"Group.1"], " ", round(missing_cohorts[min,'x']*100, 1), "%"), quote=F)
    }
  }
  if (var != "city_residence") {
     print(ggplot(terms, aes(x=.data[[var]])) +
    geom_bar() +
    labs(x=var, y="Frequency"))
  }
 
  # print statusses
  if (grepl( blacklist , var)) {
    term_vars[var,'exclude'] = T 
    print("should not be used as a predictor",quote=F)
  } else {
    term_vars[var,'exclude'] = F
  } 
  if (grepl( is_derived_from , var)) {
    term_vars[var,'use_feature'] = T 
    print("is used in feature engineering",quote=F)
  } else {
    term_vars[var,'use_feature'] = F
  } 
  # filter for variables that contain enough data, are continuous or a factor
  if(missing < 0.5 &
     (types[var] %in% c("numeric,integer") | num_unique <= 10)) {
    predictors = c(predictors, var)
    # TODO: decide if to use raw
  }
  

  print("_________________________________________",quote=F)
}
```
These are the predictors that remain when filtering for less than 50% missing, our blacklist and variables that need to be somehow transformed first: 
```{r, echo=F, comment=NA}
predictors
```
```{r}
write.csv2(term_vars, 'term_vars.csv')
```

# Student by term by course
```{r, include=F}
course_vars <- read.csv("~/EWS/Variable_request_EWS_StudentTermCourse.csv")
requested_vars = course_vars$Variable.name

courses = get_course_data(students$mellon_id)
available_vars = names(courses)
expected_available = requested_vars[requested_vars %in% available_vars]
missing_in_data = requested_vars[!requested_vars %in% available_vars]
unexpected_in_data = available_vars[!available_vars %in% requested_vars]
types = sapply(terms, class)
```

The spreadsheet knows `r length(requested_vars)` variables. The corresponding file contains `r length(available_vars)` columns. The expected variables `r paste(missing_in_data, collapse=",")` are not in the dataset. Therefore exist variables `r paste(unexpected_in_data, collapse=",")` 
```{r, include=F}
course_vars = data.frame(name=requested_vars, requested=T)
course_vars$in_data = course_vars$name %in% available_vars
extra_vars = data.frame(name=unexpected_in_data, in_data=T, requested=F)
course_vars = rbind(course_vars, extra_vars)
course_vars = column_to_rownames(course_vars, "name")
course_vars$use_raw = NA
course_vars$use_feature = NA
course_vars$exclude = NA
course_vars$num_unique = NA
course_vars$examples = NA
course_vars$type=NA
for (var in c(available_vars, unexpected_in_data)) {
  course_vars[var,'type'] = types[var]
}
course_vars$total_missing = NA
admitdates = unique(bg$admitdate)
for (a in admitdates[order(admitdates)]) {
  course_vars[paste0('missing_', a)] = NA
}
```
```{r}
# merge admitdate to courses
courses = merge(courses, bg[,c('mellon_id','admitdate')], by="mellon_id")
```

## Requested and available variables
```{r, echo=F, comment=NA}
predictors=c()
blacklist="mellon_id|group_" # should not be used as a predictor
is_derived_from="___" # is used in feature engineering
for (var in rownames(subset(course_vars, in_data))) {
  print(paste0('Variable: ', var, ", type: ", types[var], ', requested: ', if(course_vars[var,'requested'])'yes'else'no'), quote=F)
  
  unique_values = unique(courses[,var])
  num_unique = length(unique_values)
  course_vars[var,'num_unique'] = num_unique
  
  first_values = unique_values[1:min(num_unique, 5)]
  course_vars[var,'examples'] = paste0(first_values,collapse=", ")
  
  print(paste0('Values (', num_unique, ' unique): ', paste0(first_values,collapse=", "), if(num_unique > 5) ', ...' else ''),quote=FALSE)
  
  missing = mean(is.na(courses[,var]))
  course_vars[var,'total_missing'] = missing
  print(paste0("Missing: ", round(missing*100, 1), "%"), quote=F)
  
  missing_cohorts = aggregate(is.na(courses[,var]), by=list(courses$admitdate), FUN=mean)
  for (a in admitdates) {
    course_vars[var,paste0('missing_', a)] = missing_cohorts[missing_cohorts$Group.1==a,'x']
  }
  if (missing > 0.01 && missing < 0.99) {
    max = which.max(missing_cohorts[,'x'])
    min = which.min(missing_cohorts[,'x'])
    if (missing_cohorts[max,'x']-missing_cohorts[min,'x']>.5) {
      print(missing_cohorts)
    } else {
      print(paste0("Most missing: ", missing_cohorts[max,"Group.1"], " ", round(missing_cohorts[max,'x']*100, 1), "%", ", Least missing: ", missing_cohorts[min,"Group.1"], " ", round(missing_cohorts[min,'x']*100, 1), "%"), quote=F)
    }
  }
  if (var != "city_residence") {
     print(ggplot(courses, aes(x=.data[[var]])) +
    geom_bar() +
    labs(x=var, y="Frequency"))
  }
 
  # print statusses
  if (grepl( blacklist , var)) {
    course_vars[var,'exclude'] = T 
    print("should not be used as a predictor",quote=F)
  } else {
    course_vars[var,'exclude'] = F
  } 
  if (grepl( is_derived_from , var)) {
    course_vars[var,'use_feature'] = T 
    print("is used in feature engineering",quote=F)
  } else {
    course_vars[var,'use_feature'] = F
  } 
  # filter for variables that contain enough data, are continuous or a factor
  if(missing < 0.5 &
     (types[var] %in% c("numeric,integer") | num_unique <= 10)) {
    predictors = c(predictors, var)
    # TODO: decide if to use raw
  }
    print("_________________________________________",quote=F)
}
```
These are the predictors that remain when filtering for less than 50% missing, our blacklist and variables that need to be somehow transformed first: 
```{r, echo=F, comment=NA}
predictors
```

```{r}
write.csv2(course_vars, 'course_vars.csv')
```

```{r}
aggregate(current_units_completed_total ~ active_student, terms, FUN=mean)
table(terms$active_student)
aggregate(current_units_completed_total ~ full_time, terms, FUN=mean)
table(terms$full_time)

```

