---
title: "Units, based on terms and courses"
author: "Dominik Glandorf"
date: '2022-08-19'
output: pdf_document
---

```{r setup, include=FALSE}
if(!require(gridExtra)) install.packages('gridExtra')
source('read_data.R')
```

```{r prepare data}
students = get_student_sub()
bg = get_student_background_data(ids = students$mellon_id)
# get terms
terms = get_term_data(ids = students$mellon_id)
terms = merge(terms, students[,c('mellon_id','admitdate','dropout')], by="mellon_id")
# enumerate terms order by their date
terms_ordered = terms[order(terms$term_code),c('mellon_id','term_code')]
terms_ordered$term_num = ave(terms_ordered$mellon_id, terms_ordered$mellon_id, FUN=seq_along)
terms = merge(terms, terms_ordered, by=c("mellon_id","term_code"))
# get courses
courses = get_course_data(ids = students$mellon_id)
```

# Availability of units information
```{r}
units_summary_terms = is.na(terms[,names(terms)[86:121]])
colSums(units_summary_terms)[order(colSums(units_summary_terms))]
```
```{r}
units_summary_courses = is.na(courses[,names(courses)[44:47]])
colSums(units_summary_courses)[order(colSums(units_summary_courses))]
```

## Units completed total
### In term data
About how much percent of terms per user do we have units completed?
```{r}
mean(!is.na(terms$current_units_completed_total))
```
This information exists for every term of all students in term data.
```{r}
hist(terms$current_units_completed_total)
```
Hypothesis: users that have one term with very low units completed have a higher dropout chance

Some plots about current units completed:
```{r}
avg_credits_admission = aggregate(terms$current_units_completed_total, by=list("admitdate"=terms$admitdate), FUN=mean)
avg_credits_major = do.call(data.frame, aggregate(current_units_completed_total ~ major_name_1, terms, FUN = function(x) c(mean = mean(x), n = length(x))))
avg_credits_term = do.call(data.frame, aggregate(current_units_completed_total ~ term_num, terms, FUN = function(x) c(mean = mean(x), n = length(x))))
avg_credits_major_term = do.call(data.frame, aggregate(current_units_completed_total ~ major_name_1 + term_num, terms, FUN = function(x) c(mean = mean(x), n = length(x))))
```
```{r}
ggplot(avg_credits_admission, aes(x=admitdate, y=x)) +
  geom_bar(stat="identity") +
  theme_minimal() +
  labs(title = "Mean of current units completed total",
       x = "Admission",
       y = "Credit")
```
Current units does not seem to considerable vary between cohorts.

```{r}
ggplot(avg_credits_term, aes(x=term_num, y=current_units_completed_total.mean, fill=current_units_completed_total.n > 10)) +
  geom_bar(stat="identity") +
  theme_minimal() +
  labs(title = "Mean of current units completed total by term number",
       x = "Term number",
       y = "Credit",
       fill = "More than 10 terms")
```
Units in the first and 12th term are lower. If people study longer, units keep decreasing.
```{r, fig.width=14}
fill=c(rep(c("1","2"), 53))
print(avg_credits_major$major_name_1[avg_credits_major$current_units_completed_total.n<10])
ggplot(avg_credits_major, aes(x=major_name_1, y=current_units_completed_total.mean, fill=fill)) +
  geom_bar(stat="identity") +
  theme_minimal() +
  labs(title = "Mean of current units completed total per major",
       x = "Major",
       y = "Credits", fill="")+
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  scale_fill_manual(values = c("1"="#353436", "2"="#1b98e0"))+
  theme(legend.position = "none")
```
```{r, fig.height=20, fig.width=12}
ggplot(avg_credits_major_term, aes(x=term_num, y=current_units_completed_total.mean, fill=(current_units_completed_total.n > 10))) +
  geom_bar(stat="identity") +
  theme_minimal() +
  labs(title = "Mean of current units completed total per term",
       x = "Term",
       y = "Credits", fill="")+
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  theme(legend.position = "none") +
  facet_wrap(vars(major_name_1), ncol = 8)
```


### In courses data
```{r}
names(courses)
```
Let's see how much missing data we have here.
```{r}
mean(is.na(courses$units_completed))
```
This can be related to listing a lot of courses twice:
```{r}
courses[courses$mellon_id==162784,c('course_code','course_dept_code_and_num','meeting_schedule','units_completed')]
```
Why are some courses listed twice? It looks like there are different time slots for the same course.

Let's count units from courses.
```{r}
units_from_courses=aggregate(cbind(units_attempted, units_completed)~term_code+mellon_id, courses,FUN=sum)
```
And compare to term data.
```{r}
terms_merged=left_join(x=terms, y=units_from_courses, by=c("mellon_id","term_code"))
```
Do we have the information still for all the terms?
```{r}
mean(!is.na(terms_merged$units_completed))
```
Which terms are missing?
```{r}
aggregate(!is.na(units_completed)~term_code,terms_merged,FUN=mean)
```
It looks like units_completed is mostly missing in courses data from 2019 on.
How well do both data sources match apart from availability?
```{r}
mean(terms_merged$units_completed==terms_merged$current_units_completed_total, na.rm=T)
```
In 48.9% of the cases where we have both sources, they match exactly.
```{r}
mean(terms_merged$units_completed-terms_merged$current_units_completed_total, na.rm=T)
```
Mean difference is not very big in terms of the absolute values. Let's check distribution.
```{r}
hist(terms_merged$units_completed-terms_merged$current_units_completed_total, breaks=50)
```
It looks like often 0, 4 or 8 units difference is there.
```{r}
aggregate(units_completed-current_units_completed_total~admitdate, terms_merged, FUN=mean, na.action = na.omit)
```
Slight differences between cohorts.

Some example students:
```{r}
courses[courses$mellon_id==166410,c('term_desc','course_dept_code_and_num','units_attempted','units_completed')]
```
Another student:
```{r}
courses[courses$mellon_id==166128,c('term_desc','course_dept_code_and_num','units_attempted','units_completed')]
```

## Units attempted total
### In term data
```{r}
ggplot(terms, aes(x=admitdate, fill=!is.na(current_units_attempted_total))) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Current units attempted total available in term data",
       x = "Admission",
       y = "", fill="Available")
```
```{r}
length(unique(terms[!is.na(terms$current_units_attempted_total),'mellon_id']))
```
For 11612 distinct students we have information about current_units_attempted_total in term data

### In courses data
Let's see availability from courses data.
```{r}
mean(!is.na(terms_merged$units_attempted))
```
Which terms are missing?
```{r}
aggregate(!is.na(units_attempted)~term_code,terms_merged,FUN=mean)
```
It looks like units_attempted is mostly missing in courses data from 2019 on.
How well do both data sources match apart from availability?
```{r}
mean(terms_merged$units_attempted==terms_merged$current_units_attempted_total, na.rm=T)
```
In 99.4% of the cases where we have both sources, they match exactly.
```{r}
mean(terms_merged$units_attempted-terms_merged$current_units_attempted_total, na.rm=T)
```
Taking units attempted from courses data could be a good idea, given that we find this information from 2019 on.
